The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-02-25 paper/forestecology.Rmd

abstract: 
- |
  | 1. Many models for the growth of trees that incorporate the effect of interspecies competition are based on a neighborhood effect assumption whereby all trees within a fixed distance of all focal trees are considered competitors. Methods and tools are needed to quantify this competitive effect and assess the quality of all resulting models 
  | 2. We present the `forestecology` package providing methods for both 1) evaluating the effect of competitor species identity using permutation tests and 2) evaluating model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We demonstrate that both 1) competitor species identity matters and 2) that not spatially cross-validating leads to error estimates that are overly optimistic. 
  | 4. The package features 1) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, 2) functions with standardized inputs/outputs of simple features `sf` package class, and 3) R S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

@@

# knitr settings
opts_chunk$set(
  # Code output:
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  cache = TRUE,
  # Figure:
  out.width = "100%",
  fig.path = "Figures/",
  fig.width = 16 / 2.5,
  fig.height = 9 / 2.5,
  fig.align = "center",
  fig.show = "hold",
  # Etc:
  collapse = TRUE,
  comment = "##",
  # tidy = FALSE,
  tibble.print_max = 5,
  tibble.print_min = 5
)

@@

# Set output width
options(width = 70)
```

@@

$$
\newcommand{\dbh}{\text{DBH}}
\newcommand{\biomass}{\text{biomass}}
\newcommand{\BA}{\text{BA}}
y_{ij} = f(\vec{x}_{ij}) + \epsilon_{ij} = \beta_{0,j} + \beta_{\dbh,j} \cdot \dbh_{ij} + \sum_{k=1}^{K} \lambda_{jk} \cdot \BA_{ijk} + \epsilon_{ij}
$$

@@

The output `growth_scbi` is a single data frame of class `sf` that includes `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, as well a `geometry` variable encoding each tree's geolocation. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`.

The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all stems is saved in a data frame of class `sf` whereby the geolocation of each tree is presented in a `geometry` variable of type `<POINT>` and at a minimum the data contains the variables above (DAVE NOTE: THis isn't exaclty clear to me what variables above means. Maybe list all the variables needed. Do they have to have exact names?)

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, an assumption that the observations are independent underlies any cross-validation algorithm. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm by randomly resampling spatial blocks of trees @roberts_cross-validation_2017 @pohjankukka_estimating_2017. (DAVE NOTE: I don't want to change somethign I don't understand, but "randomly resampling" seems wrong to me. I don't see anything random about our block structure. Seems to me like "incorpoate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees.")

@@

DAVE NOTE::there are a couple of places where you convert to factor. Earlier with `sp` and then twice in the code block below. ANy reason we can't just do this "under the hood" in the package code? Also any reason with swtiched from undercase with underscores to `SpatialBlock_scbi` in the naming?

@@

# Associate each observation to a fold
SpatialBlock_scbi <- spatialBlock(
  speciesData = growth_scbi, k = n_fold, selection = "systematic", 
  blocks = blocks_scbi, showBlocks = FALSE, verbose = FALSE
)

growth_scbi <- growth_scbi %>%
  mutate(foldID = SpatialBlock_scbi$foldID %>% factor())
```

@@

We then compare the observed and fitted/predicted growths to compute the root mean squared error (RMSE) of our model fit. DAVE NOTE: I rounded all the RMSEs, in ecology convention to not show WAY more digits than resolution of the measurement.

```{r scbi-model-rmse}
model_rmse <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate) %>%
  round(3)
model_rmse
```

@@

These plots give the posterior distributions of parameters from equation 1 (DAVE NOTE:SOrry couldn't figure out how to make equation references work). For many pacakge users they will be of interest because they give insight into the species-specific competitive interactions. Setting `type = "intercepts"` gives posterior distributions for $\beta_{0,j}$ and `type = "dbh_slopes"` for $\beta_{dbh,j}$. These give species specific growth independent of competition. The values of more interest are plotted with `type = 'competition'` which gives the posterior distribution for $\lambda_{j,k}$ species-specific competition coefficients (i.e., the $\lambda$-matrix). Negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip trees (`litu`) have a strong negative effect on the growth of conspecifics but relatively little effect on neighbors of the other two species.

@@

focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_scbi_shuffle, newdata = focal_vs_comp_scbi))

model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate) %>%
  round(3)
model_rmse_shuffle
```

@@

To this end, we use the `foldID` variable defined in Section \ref{spatial-information} whereby all focal trees are assigned to one of `r n_fold` spatially contiguous blocks that act as folds in our cross-validation routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic} presents a schematic illustrating this scheme for fold 1 (bottom-left) as the test set and folds 2, 3, and 4 as the training sets. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set to compute fitted/predicted values, and compute the RMSE of the observed versus predicted growths. We repeat this procedure 3 more times with each of the three remaining folds acting as the test set and then average all four resulting RMSE's. Furthermore, in order to maintain spatial independence between the test and training set, a fold buffer that extend outwards from the boundary of the test set is computed; all trees falling within this fold buffer are excluded from the training set. DAVE NOTE: I repalced "buffer" with "fold buffer" just to keep clear that this is different than our global buffer. But if you think that distinction is not necessary feel free to switch back!

@@

```{r scbi-spatial-cv}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)
```
```{r}
model_rmse_cv <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate) %>%
  round(3)
model_rmse_cv
```

>>>>

abstract: 
- |
  | 1. Many models for the growth of trees that incorporate interspecies competition are based on a neighborhood effect assumption whereby all trees within a fixed distance of a focal tree are considered competitors. Methods and tools are needed to quantify this competitive effect and assess the quality of all resulting models. 
  | 2. We present the `forestecology` package providing methods for both 1) evaluating the effect of competitor species identity using permutation tests and 2) evaluating model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We demonstrate that both 1) competitor species identity matters and 2) that not spatially cross-validating leads to error estimates that are overly optimistic. 
  | 4. The package features 1) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, 2) functions with standardized inputs/outputs of simple features `sf` package class, and 3) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

@@

# knitr settings
opts_chunk$set(
  # Code output:
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  cache = TRUE,
  # Figure:
  out.width = "100%",
  fig.path = "Figures/",
  fig.width = 16 / 2.5,
  fig.height = 9 / 2.5,
  fig.align = "center",
  fig.show = "hold",
  # Etc:
  collapse = TRUE,
  comment = "##"
)

@@

# Set output width
options(
  width = 70, 
  # Decimal precision
  digits = 3, 
  # Number of rows for tibble printing
  tibble.print_max = 5, tibble.print_min = 5
)
```

@@

\begin{equation}
\label{eq:model}
y_{ij} = f(\vec{x}_{ij}) + \epsilon_{ij} = \beta_{0,j} + \beta_{\text{dbh},j} \cdot \text{dbh}_{ij} + \sum_{k=1}^{K} \lambda_{jk} \cdot \text{BA}_{ijk} + \epsilon_{ij}
\end{equation}

@@

The output `growth_scbi` is a single data frame of class `sf` that includes variables `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, and `geometry`, the `sf` package's encoding of geolocations of type `<POINT>`. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`. The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all stems is saved in a data frame of class `sf` and at a minimum the data contains the variables above. TODO: enumerate what these variables are and say if they need exact names

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, an assumption that the observations are independent underlies any cross-validation algorithm. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees. @roberts_cross-validation_2017 @pohjankukka_estimating_2017.

@@

@@

# Associate each observation to a fold
spatial_block_scbi <- spatialBlock(
  speciesData = growth_scbi, k = n_fold, selection = "systematic", 
  blocks = blocks_scbi, showBlocks = FALSE, verbose = FALSE
)

growth_scbi <- growth_scbi %>%
  mutate(foldID = spatial_block_scbi$foldID %>% factor())
```

@@

We then compare the observed and fitted/predicted growths to compute the root mean squared error (RMSE) of our model fit.

```{r scbi-model-rmse}
model_rmse <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
model_rmse
```

@@

These plots give the posterior distributions of parameters from Equation \ref{eq:model}. For many pacakge users they will be of interest because they give insight into the species-specific competitive interactions. Setting `type = "intercepts"` gives posterior distributions for $\beta_{0,j}$ and `type = "dbh_slopes"` for $\beta_{dbh,j}$. These give species specific growth independent of competition. The values of more interest are plotted with `type = 'competition'` which gives the posterior distribution for $\lambda_{j,k}$ species-specific competition coefficients (i.e., the $\lambda$-matrix). Negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip trees (`litu`) have a strong negative effect on the growth of conspecifics but relatively little effect on neighbors of the other two species.

@@

focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_scbi_shuffle, newdata = focal_vs_comp_scbi))
```
```{r}
model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
model_rmse_shuffle
```

@@

To this end, we use the `foldID` variable defined in Section \ref{spatial-information} whereby all focal trees are assigned to one of `r n_fold` spatially contiguous blocks that act as folds in our cross-validation routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic} presents a schematic illustrating this scheme for fold 1 (bottom-left) as the test set and folds 2, 3, and 4 as the training sets. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set to compute fitted/predicted values, and compute the RMSE of the observed versus predicted growths. We repeat this procedure 3 more times with each of the three remaining folds acting as the test set and then average all four resulting RMSE's. Furthermore, in order to maintain spatial independence between the test and training set, a fold buffer that extend outwards from the boundary of the test set is computed; all trees falling within this fold buffer are excluded from the training set.

@@

```{r scbi-spatial-cv}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)
```
```{r}
model_rmse_cv <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
model_rmse_cv
```
